---
title: "UnitTestingHW"
author: "Laura Tomkins"
date: "March 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, include = TRUE)
```

# Overview of project

The main goal of this project is to explore the environmental condtions of clearing events in Stratocumulus clouds over the Southeast Atlantic Ocean. This project will download reanalysis data from online, read in this data, find the correct fields to plot, and then plot said fields. Possible variables that will be examined includes pressure, potential temperature, wind, etc. 

# Top most design

The first function `get_data` will download reanalysis data from the web given desired dates for data and the general source for the data. Probable syntax: `data <- get_data(starttime, endtime, source)`. A test will be completed by checking if directory data is stored is empty or non-empty

```{r get_data test}
data <- get_data(starttime, endtime, source)
maindir <- "data"
subdir <- starttime
if (dir.exists(file.path(maindir, subdir))){
  paste0('get_data test passed')
} else{
  pasteo('get_data test failed')
}
```

The second function will be `get_variable` which will isolate the data for a given variable. Probable syntax: `var_data <- get_variable(data_file, variable)`. A test will be completed to check that the data frame is non-empty and has data which makes sense.

```{r get_variable test}
var_data <- get_variable(data, "PRES")
if (exists(var_data)){
  paste0('get_variable test passed')
} else{
  pasteo('get_variable test failed')
}
```

The third function will be `plot_variable` which will plot the variables obtained in `get_variable`. Probable syntax: `plot <- plot_variable(variable, latitude_lims, longitude_lims)`. Most of the plots are likly to be spatial, hence the usage of latitude and longitude limits. Unsure of ways to test as present moment.

```{r plot_variable test}
plot <- plot_variable(variable, latitude_lims, longitude_lims)
if (exits(plot)){
  paste0('plot_variable test passsed')
} else{
  pasteo('plot_variable test failed')
}
```

More functions are likely to be added in the process and current functions are likely to be edited/re-designed.

