---
title: "Temperature and Precipitation Data"
author: "Laura Tomkins"
date: "March 11, 2018"
output: html_document
#bibliography: EVRN_assignment1.bib
#csl: science.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

```{r, include = FALSE}
# Data Chunk
provided <- TRUE
nrecords <- 40

if (provided==TRUE){
  precip <- readRDS("USAAnnualPcpn1950_2008.rds")
  temp <- readRDS("USAAnnualTemp1950_2008.rds")
} else {
  tmep <- readRDS("USAAnnualTemp1950_2016.rds")
  # Insert code for getting data online
}
```

## Introduction
It is fairly well accepted among the scientific community that the global temperatures around the globe are increasing. Temperature records have shown increasing temperatures over the past century compared to pre-industrial records. This is increase is largely anthropogenic, due to the increase in greenhouse gases from burning fossil fuels. This report will attempt to use some temperature records available to show these changing temperatures. Along with global warming, increasing the concentration of CO2 also has implications for other parts of our atmosphere. Climate change will likely change our weather patterns and thus our precipitation patterns. Precipitation data is also provided in this report which will also be analyzed for evidence of climate change. [@RN1;@RN5]

#### Research Questions

For this assignement we will attempt to answer the following 3 questions provided from the assignment instructions:  
1. Can we see climate warming for ourselves by analyzing these data?  
2. In what parts of the country are temperatures getting warmer? Are there any parts that
actually got colder over 1950-2010?  
3. Can we say, based on these data, What change has occurred in precipitation over the period?  

#### Data
The data used in this analysis is annual temperature and precipitation data from various locations over the continental U.S. Data was either provided from the instructor or downloaded from online. It should be noted that this data hasbeen quality controlled to remove stations which have less than `r nrecords` years of data.

```{r, include = FALSE}
# Cleaning chunk
source('data_cleaner.R')

cleanprecip <- data_cleaner(precip,nrecords,provided)
cleantemp <- data_cleaner(temp,nrecords,provided)

remove(temp)
remove(precip)
```

## Temperature Data

My first approach is to plot the temperature data from 1950 (which is shown in Figure 1) and then to plot the temperature data from 2008 (Figure 2) to see if there are any visual patterns which stand out and could indicate that 2008 is warmer than 1950. Looking at figures 1 and 2 there doesn't seem to be any strong signal either way. This is not a great method for looking at patterns in climate. To examine patterns in climate we ideally need data over multiple years as individual years aren't sufficient. For example, while we should expect 2008 to be a warmer year than 1950, this might not be the case as there are other factors besides climate change which affect the annual average temperature over the United States (ENSO, for example). 

```{r, include=FALSE}
# Plotting Chunk
library(ggplot2)
library(ggthemes)
library(ggmap)

conus <- get_stamenmap(c(left=-125, bottom=25.75, right=-67, top=49), zoom = 5, maptype = 'toner-lite')
```

```{r 1950temp, fig.cap = 'Figure 1. 1950 Temperature Data'}
gg <- ggmap(conus) + 
  geom_point(data=cleantemp, aes(x=lon,y=lat,color=`1950`),size=1,na.rm=TRUE)+
  labs(title = '1950 Temperature Data', x = 'Longitude', y = 'Latitude') + 
  scale_color_gradientn(colors=rev(rainbow(7)), limits=c(10,80), na.value='white') 
gg
```

```{r 2008temp, fig.cap = 'figure 2. 2008 Temperature Data'}
gg <- ggmap(conus) + 
  geom_point(data=cleantemp, aes(x=lon,y=lat,color=`2008`),size=1,na.rm=TRUE)+
  labs(title = '2008 Temperature Data', x = 'Longitude', y = 'Latitude') + 
  scale_color_gradientn(colors=rev(rainbow(7)), limits=c(10,80), na.value='white') 
gg
```

Instead of just comparing 2 years I will compare the average of the first half of the data to the average of the second half of the data and see if any patterns emerge here.

```{r}
# Compare average of first and second half

mididx = round((NCOL(cleantemp)-3)/2)
firsthalf = cleantemp[,3:(mididx+3)]
secondhalf = cleantemp[,(mididx+3):NCOL(cleantemp)]
first3 = cleantemp[,4:6]
last3 = cleantemp[,(NCOL(cleantemp)-2):NCOL(cleantemp)]

avg1 = rowMeans(firsthalf, na.rm=TRUE)
avg2 = rowMeans(secondhalf, na.rm=TRUE)
af = rowMeans(first3, na.rm=TRUE)
al = rowMeans(last3, na.rm=TRUE)

avgdat <- cbind(cleantemp,avg1,avg2,(avg2-avg1),af,al,(al-af))
```

```{r firstavg, fig.cap = 'Figure 3. Average of first half of Temperature Data'}
gg <- ggmap(conus) + 
  geom_point(data=avgdat, aes(x=lon,y=lat,color=`avg1`),size=1,na.rm=TRUE)+
  labs(title = 'First half average', x = 'Longitude', y = 'Latitude') + 
  scale_color_gradientn(colors=rev(rainbow(7)), limits=c(10,80),na.value='white') 
gg
```

```{r secondavg, fig.cap = 'Figure 4. Average of second half of Temperature Data'}
gg <- ggmap(conus) + 
  geom_point(data=avgdat, aes(x=lon,y=lat,color=`avg2`),size=1,na.rm=TRUE)+
  labs(title = 'Second half average', x = 'Longitude', y = 'Latitude') + 
  scale_color_gradientn(colors=rev(rainbow(7)), limits=c(10,80),na.value='white') 
gg
```

Looking at figures 3 and 4, it appears as though the average of the second half of the data shows generally warmer temperatures, which we would expect in a warming climate. Figure 5 below shows the difference between the average of the first and second half of the data (second minus first). Positive values (red) would indicate warmer averages in the second half of the data and vice versa for negative values (blue). Here we see that the data over the map is overwhelmingly positive indicating that the second half of the data is warmer than the first half.

```{r avgdiff, fig.cap = 'Figure 5. Difference between first and second half average of temperature data (second-first)'}
gg <- ggmap(conus) + 
  geom_point(data=avgdat, aes(x=lon,y=lat,color=`(avg2 - avg1)`),size=1,na.rm=TRUE)+
  labs(title = 'Second half average - First half average', x = 'Longitude', y = 'Latitude') + 
  scale_color_distiller(palette='RdBu', limits=c(-15,15),na.value='white') 
gg
```

```{r avgplot, fig.camp = 'Figure 6. Annual average annual temperature for U.S.'}
years = as.numeric(colnames(cleantemp)[4:NCOL(cleantemp)])
avgbyyear = colMeans(cleantemp[4:NCOL(cleantemp)], na.rm=TRUE)
plot(years, avgbyyear, main='Average Temperature for U.S.', pch=16, xlab='Years', ylab='Temperature', col='dark blue')
lines(lowess(years,avgbyyear), col="dark red") # lowess line (x,y)
fit <- lm(avgbyyear ~ years)
abline(coef(fit)[1:2], col='blue')
cf <- round(coef(fit), 2) 
eq <- paste0("Temperature = ", cf[1],
             ifelse(sign(cf[2])==1, " + ", " - "), abs(cf[2]), " * Year ")
r2 <- paste0('R-squared = ',round(summary(fit)$r.squared,4))
mtext(eq, 3, line=-2)
mtext(r2, 3, line=-3)
```

Figure 6 shows a plot of the average annual temperature for the U.S. Years are shown on the x axis and average temperature on the y axis. The blue line represents the regression line with the equation and r-squared value shown at the top and the red line represents the locally weighted scatterplot smoothing (LOWESS) line. We can see that the regression line has a general positive trend, however the r-squared value is very small (r-squared = `r round(summary(fit)$r.squared,4)`)and the data do not fit this line particularly well. The LOWESS line shows a trend in the data which is more curved as there appearedto be some cooling in the U.S. during the 1970s followed by warming into the 1980s through the 2000s.

<!--I've also compared the average of the first 3 years of data and the last 3 years of data which are shown below in Figure 6. 

```{r avg3, fig.cap = 'Figure 6. Difference between first 3 years and last 3 years of temperature data (last-first)'}
gg <- ggmap(conus) + 
  geom_point(data=avgdat, aes(x=lon,y=lat,color=`(al - af)`),size=1,na.rm=TRUE)+
  labs(title = 'Last 3 years average - First 3 years average', x = 'Longitude', y = 'Latitude') + 
  scale_color_distiller(palette='RdBu', limits=c(-15,15),na.value='white') 
gg
``` 
-->

## Precipitation

The next part of this analysis will focus on the change in precipitation patterns. The first thing we should notice as we look at figures 6 and 7 is that we oonly have precipitation data for Ohio, Tennessee, and Alabama. This present issues for making conclusions about patterns over the whole US since we don't have data for most of our area. We can only make conclusions about those states where we have data. 

```{r 1950precip, fig.cap = 'Figure 6. 1950 Precipitation Data'}
gg <- ggmap(conus) + 
  geom_point(data=cleanprecip, aes(x=lon,y=lat,color=`1950`),size=1,na.rm=TRUE)+
  labs(title = '1950 Precipitation Data', x = 'Longitude', y = 'Latitude') + 
  scale_color_gradientn(colors=rev(rainbow(7)), limits=c(20,65), na.value='white') 
gg
```

```{r 2008precip, fig.cap = 'Figure 7. 2008 Precipitation Data'}
gg <- ggmap(conus) + 
  geom_point(data=cleanprecip, aes(x=lon,y=lat,color=`2008`),size=1,na.rm=TRUE)+
  labs(title = '2008 Precipitation Data', x = 'Longitude', y = 'Latitude') + 
  scale_color_gradientn(colors=rev(rainbow(7)), limits=c(20,65), na.value='white') 
gg
```

```{r}
# Compare average of first and second half

mididxp = round((NCOL(cleanprecip)-3)/2)
firsthalfp = cleanprecip[,3:(mididxp+3)]
secondhalfp = cleanprecip[,(mididxp+3):NCOL(cleanprecip)]
#first3p = cleanprecip[,4:6]
#last3p = cleanprecip[,(NCOL(cleanprecip)-2):NCOL(cleanprecip)]

avg1p = rowMeans(firsthalfp, na.rm=TRUE)
avg2p = rowMeans(secondhalfp, na.rm=TRUE)
#af = rowMeans(first3p, na.rm=TRUE)
#al = rowMeans(last3p, na.rm=TRUE)

avgdatp <- cbind(cleanprecip,avg1p,avg2p,(avg2p-avg1p))#,af,al,(al-af))
```

```{r avgdiffprecip, fig.cap = 'Figure 8. Difference between first and second half average of precipitation data (second-first)'}
gg <- ggmap(conus) + 
  geom_point(data=avgdatp, aes(x=lon,y=lat,color=`(avg2p - avg1p)`),size=1,na.rm=TRUE)+
  labs(title = 'Second half average - First half average', x = 'Longitude', y = 'Latitude') + 
  scale_color_distiller(palette='BrBG', direction=1, limits=c(-10,10),na.value='white') 
gg
```

Figure 8 shows the difference between the average of the first half and second half of precipitation data. Brown indicates negative values, or areas where the precipitation has decreased for the second half of the data and blue indicates positive values, or areas where the precipitation has increased for the second half of the data. For the states where we have data we can see mostly blue values which indicates that precipitation has increased over these areas.

