---
title: "Assignment 2: Temperature and Precipitation Patterns"
author: "Laura Tomkins"
date: "March 2018"
output: html_document
bibliography: EVRN_assignment2.bib

# Modularity flag format: #! Modularity Lesson #X: blah blah blah
---

<!--***DAN: Fine comments by searching "***DAN"-->

<!--***DAN: Excellent use of git - plenty of commits, sensible commit messages, and no files committed that should not be. My one criticism is, you do not have to add the date to your commit message because date and time info is automatically stored with the commit, as is information about which files were changed and who did the commit.-->

<!--***DAN: The write up is certainly sufficent for this assignment. Nice plots, esp. fig 2, which is quite smilar to what I got originally with the provided data, and better presented than my plot. Nice use of references. No problems here.-->

<!--***DAN: You also did a very good job with modularity! For the most part, the code is very clean because of that! 10/10-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

```{r data_setup, include = FALSE}
# Data Chunk
# ! Modularity Task Lesson #1: defining constants
provided <- TRUE  # DO NOT CHANGE TO FALSE UNLESS YOU WANT TO SIT AROUND FOR A COUPLE HOURS
nrecords <- 40

if (provided==TRUE){
  precip <- readRDS("USAAnnualPcpn1950_2008.rds")
  temp <- readRDS("USAAnnualTemp1950_2008.rds")
} else {
  source('data_download.R')
  temp <- readRDS("USAAnnualTemp1950_2016.rds")
  tryCatch(temp <- data_download('19500101','20101231',state.abb))
  precip <- readRDS("USAAnnualPcpn1950_2008.rds")
} # ! Modularity Lesson #2: extensible code (for the future easy to obtain up to date data)
```

## Introduction
It is fairly well accepted among the scientific
<!--***DAN: I think it is fair to say something considerably stronger than this! Like "very well accepted", for instance.-->
community that the global temperatures around the globe are increasing. Temperature records have shown increasing temperatures over the past century compared to pre-industrial records. This trend has been documented as early as the 1960s [@mitchell1961recent]. This is increase is thought to be largely anthropogenic, due to the increase in greenhouse gases from burning fossil fuels. This report will attempt to use some temperature records available to show these changing temperatures. Along with global warming, increasing the concentration of CO2 also has implications for other parts of our atmosphere. Climate change will likely change our weather patterns and thus our precipitation patterns [@IPCCSummaryStocker2013]. Precipitation data is also provided in this report which will also be analyzed for evidence of climate change.


## Methods
#### Research Questions
For this assignement we will attempt to answer the following 3 questions provided from the assignment instructions:  
1. Can we see climate warming for ourselves by analyzing these data?  
2. In what parts of the country are temperatures getting warmer? Are there any parts that
actually got colder over 1950-2010?  
3. Can we say, based on these data, What change has occurred in precipitation over the period?  

To analyze trends in the data, a regression analysis will be used to look at any general trends in the averages of the data. To analyze spatial trends in the data, plots of regression slopes for each location will be plotted to see areas which increase and areas which decrease.

<!--***DAN: Great modularity here.-->
```{r data_clean, include = FALSE}
# Cleaning chunk
source('data_cleaner.R')

cleanprecip <- data_cleaner(precip,nrecords,provided=TRUE)
cleantemp <- data_cleaner(temp,nrecords,provided)
# ! Modularity Lesson # 6: writing scripts to avoid copying code
# ! Modularity Lesson # 7: psuedocode (inside data_cleaner function)

remove(temp)
remove(precip)
```

## Data
The data used in this analysis is annual temperature and precipitation data from various locations over the continental U.S. Data was either provided from the instructor or downloaded from online. It should be noted that this data has been quality controlled to remove stations which have less than `r nrecords` years of data. The temperature data consists of observations from `r length(cleantemp$name)` stations and the precipitation data consists of observations from `r length(cleanprecip$name)` stations. The termperature data is fairly well distributed over the entire continental U.S. however, the precipitation data only covers 3 states which may result in some complications.

## Results

### Temperature
The following is an analysis ofthe temperature data.

```{r plot_setup, include=FALSE}
# Plotting Chunk
library(ggplot2)
library(ggthemes)
library(ggmap)

conus <- get_stamenmap(c(left=-125, bottom=25.75, right=-67, top=49), zoom = 5, maptype = 'toner-lite')
```

```{r avgplot, fig.cap = 'Figure 1. Annual average annual temperature for U.S.'}
years = as.numeric(colnames(cleantemp)[4:NCOL(cleantemp)])
avgbyyear = colMeans(cleantemp[4:NCOL(cleantemp)], na.rm=TRUE)
plot(years, avgbyyear, main='Average Temperature for U.S.', pch=16, xlab='Years', ylab='Temperature', col='dark blue')
lines(lowess(years,avgbyyear), col="dark red") # lowess line (x,y)
fit <- lm(avgbyyear ~ years)
abline(coef(fit)[1:2], col='blue')
cf <- round(coef(fit), 2) 
eq <- paste0("Temperature = ", cf[1],
             ifelse(sign(cf[2])==1, " + ", " - "), abs(cf[2]), " * Year ")
r2 <- paste0('R-squared = ',round(summary(fit)$r.squared,4))
mtext(eq, 3, line=-2)
mtext(r2, 3, line=-3)
```

```{r trendplot, fig.cap = 'Figure 2. Spatial trends for average annual temperature for U.S.'}
# Dependencies: need to have spatial plot set up (conus) and data cleaned (cleantemp)
#! Modularity lesson #8: be aware of dependencies
source('data_trend.R')

trends <- data_trend(cleantemp, years)
trenddat <- cbind(cleantemp,trends)
gg <- ggmap(conus) + 
  geom_point(data=trenddat, aes(x=lon,y=lat,color=trends),size=1,na.rm=TRUE)+
  labs(title = 'Slope of regression line for U.S. Temperature data', x = 'Longitude', y = 'Latitude') + 
  scale_color_distiller(palette='RdBu', limits=c(-10,10),na.value='white') 
gg
```

Figure 1 shows a plot of the average annual temperature for the United States. Years are shown on the x axis and average temperature on the y axis. The blue line represents the regression line with the equation and r-squared value shown at the top and the red line represents the locally weighted scatterplot smoothing (LOWESS) line. We can see that the regression line has a general positive trend, however the r-squared value is very small (r-squared = `r round(summary(fit)$r.squared,4)`) and the data do not fit this line particularly well. The LOWESS line shows a trend in the data which is more curved as there appeared to be some cooling in the U.S. during the 1970s followed by warming into the 1980s through the 2000s. Because this averages temperatures for all the locations it does not account of areas where there may be cooling and areas where there may be warming, so this is not necessarily the best way to see this. We are also slightly limited because if we want to study the warming trend we should really look at temperature before the industrial revolution to get a better sense of the warming.

To observe a spatial trend of the data the slope of the regression line for each station is plotted in figure 2. Here we can see a mixture of cooling (blue dots) and warming(red dots). It appears as though there is a concentration of blue dots around the southern plains/deep south region of the U.S. Most other locations appear to show warming trends.

### Precipitation

The next part of this analysis will focus on the change in precipitation patterns. As is stated previously, we only have precipitation data for Ohio, Tennessee, and Alabama. This present issues for making conclusions about patterns over the whole US since we don't have data for most of our area. We can only make conclusions about those states where we have data. 

```{r avgplotp, fig.cap = 'Figure 3. Annual average annual precipitation for U.S.'}
yearsp = as.numeric(colnames(cleanprecip)[4:NCOL(cleanprecip)])
avgbyyearp = colMeans(cleanprecip[4:NCOL(cleanprecip)], na.rm=TRUE)
plot(yearsp, avgbyyearp, main='Average Precipitation for U.S.', pch=16, xlab='Years', ylab='Precipitation', col='dark blue')
lines(lowess(yearsp,avgbyyearp), col="dark red") # lowess line (x,y)
fitp <- lm(avgbyyearp ~ yearsp)
abline(coef(fitp)[1:2], col='blue')
cfp <- round(coef(fitp), 2) 
eqp <- paste0("Precipitation = ", cfp[1],
             ifelse(sign(cfp[2])==1, " + ", " - "), abs(cfp[2]), " * Year ")
r2p <- paste0('R-squared = ',round(summary(fit)$r.squared,4))
mtext(eqp, 3, line=-2)
mtext(r2p, 3, line=-3)
```

```{r trendplotp, fig.cap = 'Figure 4. Spatial trends for average annual precipitation for U.S.'}
source('data_trend.R')

trendsp <- data_trend(cleanprecip, years)
trenddatp <- cbind(cleanprecip,trendsp)
gg <- ggmap(conus) + 
  geom_point(data=trenddatp, aes(x=lon,y=lat,color=trendsp),size=1,na.rm=TRUE)+
  labs(title = 'Slope of regression for U.S. Precipitation data', x = 'Longitude', y = 'Latitude') + 
  scale_color_distiller(palette='BrBG', direction=1,limits=c(-1,1),na.value='white') 
gg
```

Similar to figures 1 and 2, figures 3 and 4 show the same thing but for the precipitation data. Figure 3 shows a general positive trend, however the data do not fit the line very well and the r-squared value is only `r round(summary(fitp)$r.squared,4)` which is not great. Again, this is probably not very significant anyway since we only have data for 3 states.

Figure 4 shows the regression slope for the individual locations. Here we find that most locations show an increase in precipitation, however there are a few locations which show a decrease in precipitation.



## References